
\begin{titlepage}
\rightline{\LARGE \textbf{Mateusz Jadczak}}

\vspace*{60mm}
\begin{center}
\Huge
\textbf{An implementation of Google Dataflow in the Elixir programming language} \\[5mm]
Computer Science Tripos -- Part II \\[5mm]
Robinson College \\[5mm]
19 May 2017 \\[20mm]
\end{center}
\end{titlepage}

% Proforma


{\let\cleardoublepage\clearpage \chapter*{Proforma} % Start on next side

{\large\tabulinesep=1mm
\begin{tabu}{lX[p]}
Name:               & \textbf {Mateusz Jadczak}                       \\
College:            & \textbf {Robinson}                     \\
Project Title:      & {\setstretch{0.8}\textbf{An implementation of Google Dataflow in the Elixir programming language}} \\
Examination:        & \textbf {Computer Science Tripos -- Part II, June 2017}  \\
Word Count:         & $\sim$11,700\footnotemark[1]  \\
Project Originators: & Dr Alastair Beresford \& Dr Martin Kleppmann     \\
Supervisor:         &   Dr Alastair Beresford           \\ 
\end{tabu}
}
\footnotetext[1]{This word count was computed
using \texttt{texcount -1 -merge -sum mnj24-dissertation.tex}.
}
\stepcounter{footnote}

\bigskip

\section*{Original aims of the project}

To implement a Domain Specific Language (DSL) for the construction of Apache Beam-compatible Pipelines, as well as an interface to the Java codebase allowing for the execution of these Pipelines using existing Beam software.

\section*{Work completed}

A DSL was implemented which requires significantly less code to express Model concepts than the Java API.
It also adopts standard Elixir idioms.
A full Beam-like runner was implemented in Elixir, instead of interfacing with a Java runner.

The runner implemented can execute computation Pipelines expressed in the DSL and supports a large number of core Beam features useful in common applications.
It displays superior performance under load in terms of resource usage, throughput and latency when compared against both the local Beam Java runner and the production-optimised Apache Flink runner.
An experiment using a replicated Twitter API stream shows that the implementation is capable of throughput of over \num3$\times$ the entire Twitter Firehose.

\section*{Special difficulties}

None.

%\newpage
\section*{Declaration}

I, Mateusz Jadczak of Robinson College, being a candidate for Part II of the Computer Science Tripos, hereby declare that this dissertation and the work described in it are my own work, unaided except as may be specified below, and that the dissertation does not contain material that has already been used to any substantial extent for a comparable purpose. 

\bigskip
\leftline{Signed}

\medskip
\leftline{Date}


\tableofcontents
\listoffigures
\listoflistings
\listoftables
} % start opening on right side only
%\newpage
